---
title: 'Demo: Leaflet Maps'
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---

## Making a map with points in R with leaflet

We're going to be doing a lot of the stuff from the RStudio [leaflet tutorial](https://rstudio.github.io/leaflet/). This walkthrough is a variation of [Amelia McNamara](https://github.com/AmeliaMN)'s version.

Load `leaflet` first:
```{r}
library(leaflet)
```

Let's start with a simple map. We create a map widget by calling `leaflet()`, and then add layers (ex. `addTiles`, `addMarkers`, `addPolygons`) to modify the widget. 
```{r}

m <- leaflet() %>%
  setView(-72.518978, 42.381050, zoom = 10) %>%
  addTiles() %>%  # Load the default OpenStreetMap tile set
  addMarkers(lng=-72.6403, lat=42.3163, popup = "Smith!")
m

```

Note that the syntax here is similar to that of `ggplot2`. Please consult the [Leaflet documentation](http://rstudio.github.io/leaflet/) for more details.

## Okay, let's do something with actual data

Lets look at [storm data](http://www.ncdc.noaa.gov/stormevents/ftp.jsp) from the NOAA. It comes in a few files that we need to join together in order to use. For convenience, I've put them on the course website. You can download them manually, as .zip files:

```{r}
library(readr)
library(dplyr)

# One file contains information about the storm
# ignore the warning 
stormdetails <- read_csv("too-big-for-git/StormEvents_details-ftp_v1.0_d2023_c20240917.csv")

# The other contains actual location data
stormlocs <- read_csv("too-big-for-git/StormEvents_locations-ftp_v1.0_d2023_c20240917.csv")

# We'll want to use them together, so we'll use a join
storms <- stormlocs %>%
  left_join(stormdetails, by="EVENT_ID")
```
```{r}
glimpse(storms)
```

## Mapping lightning strikes
Let's pull out the lightning strikes and map them as part of a `dplyr` chain
```{r}
lightning_map <- storms %>%               # Start with the storm data
  filter(EVENT_TYPE == "Lightning") %>%   # Filter down to just the lightning events
  leaflet() %>%  
    addTiles() %>%  
    addMarkers(~LONGITUDE, ~LATITUDE, popup = "Zap!")  

lightning_map
```

## Challenge: 
- Find another storm type to map
- Bonus-- add popups!

## One approach

```{r}
mtw_map <- storms %>%
  filter(EVENT_TYPE == "Marine Thunderstorm Wind") %>%
  leaflet() %>%
    addTiles() %>% 
    addMarkers(~LONGITUDE, ~LATITUDE, popup = ~EVENT_NARRATIVE)

mtw_map
```


## Polygons

First let's grab some data (and do a little conversion)

```{r}
tornados <- storms %>%
  filter(EVENT_TYPE=="Tornado") %>%
  mutate(DAMAGE_PROPERTY = as.numeric(sub("K", "", DAMAGE_PROPERTY, fixed = TRUE)))
```

Then let's start with something easy(ish) -- circles
```{r}
m <- leaflet(data = tornados) %>%
  addTiles() %>% 
  addCircles(~LONGITUDE, ~LATITUDE, 
             weight = 1, 
             radius = ~DAMAGE_PROPERTY*100, # Map the radius of the circle to amount of damage
             popup = ~EVENT_NARRATIVE)      # Include details about the tornado
m
```

## Polgyons come in shapefiles
Most boundaries (state, national, etc) are provided in terms of polygons. Major mapping software ArcGIS, from ESRI, has essentially set the standard formats. There are many files with different extensions: .prj (the projection), .shp (the shapefile), .cpg (??), .dbf (??), .shx (??). 

You need special software or packages to work with shapefiles. 

## State shapefiles

This portion of the tutorial is based on: [R Graph Gallery](https://r-graph-gallery.com/168-load-a-shape-file-into-r.html)

I got the shapefiles used below from [the Census](https://www.census.gov/geo/maps-data/data/cbf/cbf_state.html). You can choose the resolution. If you want, the zipfile of the shapes I used is on the course website (cb_2018_us_state_500k.zip).  

We're going to use the `sf` package to deal with shapefiles. The sf package offers the `st_read()` and the `read_sf()` functions that allow us to read shapefile using the following syntax.

As a result you get a geospatial object (`my_sf` here) that contains all the information we need for further mapping.
```{r}
library(sf)

my_sf <- read_sf("too-big-for-git/cb_2018_us_state_500k/cb_2018_us_state_500k.shp")

print(my_sf) #tells you the max and min coordinates, the kind of projection in use and the the first few rows of the data associated with the regions.

nrow(my_sf) #how many regions you have

```
Notice how each region has a POLYGON associated with it. 

## Mapping shapefiles

We can map shapefiles using `ggplot2` and the `geom_sf()` function:  

```{r}
library(ggplot2)

ggplot(my_sf) +
  geom_sf(fill = "#69b3a2", color = "white") + 
  theme_void()

```

What if we want just a specific region?

```{r}

newEngland <- my_sf[(my_sf$STUSPS == "ME" | my_sf$STUSPS == "NH" | my_sf$STUSPS == "VT" | 
                      my_sf$STUSPS == "MA" | my_sf$STUSPS == "RI" | my_sf$STUSPS == "CT"), ]

plot(st_geometry(newEngland),
  col = "steelblue", lwd = 0.5
)

```

## Your Turn!

Put it all together! See if you can make a map of your own data, and embed it within a dashboard. **Bonus:** can you make them interact with one another?

## Resources
Follow the links above to more resources for geospatial visualization with R

