---
title: 'Demo: Plotly'
output:
  html_document:
    css: lab.css
    highlight: pygments
    theme: cerulean
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

This tutorial comes from Carson Sievert's [Plotly for R Master Class](https://plotly-book.cpsievert.me/index.html).

# Case Study: housing sales in Texas

## Loading `plotly`
The `plotly` package depends on `ggplot2` which bundles a data set on monthly housing sales in Texan cities acquired from the [TAMU](http://recenter.tamu.edu/) real estate center. After the loading the package, the data is "lazily loaded"" into your session, so you may reference it by name:

```{r}
library(plotly)
txhousing
```

## Good old `ggplot2`
Let's see if there's any pattern in house price behavior over time:
```{r}
p <- txhousing %>%
  group_by(city) %>%
    ggplot(aes(x = date, y = median)) +
      geom_line(aes(group = city), alpha = 0.2)
p
```

## Make it interactive
It'd be nice if we could see which city each line corresponds to when we hover. `plotly` makes this easy! Just wrap your `ggplot` object in the  `ggplotly()` function:

```{r}
class(p)
ggplotly(p)
```

If we just want the city name, we can specify exactly what to put in the tooltip:
```{r}

ggplotly(p, tooltip = "city")

```

## Tidying up with `plot_ly()`
We can also build `plotly` objects directly using the `plot_ly()` function along with `dplyr`-like syntax. Why would we want to? Well, for one thing, `plot_ly()` recognizes and preserves groupings created with `dplyr`’s `group_by()` function:

```{r}
library(dplyr)
tx_grouped <- group_by(txhousing, city)

# initiate a plotly object with date on x and median on y
p <- plot_ly(tx_grouped, x = ~date, y = ~median)

plotly_data(p)
```

Since we didn't specify any mapping, the plot defaults to a scatterplot:
```{r}
p
```

Let's change that to a line chart. Similar to `geom_line()` in `ggplot2`, the `add_lines()` function connects (a group of) x/y pairs with lines in the order of their x values and returns the transformed `plotly` object:

```{r}
p %>%
  add_lines(alpha = 0.2, name = "Texan Cities")
```

## Highlighting
Want to highlight a particular line? Filtering works, and since each `add_lines()` call returns a pointer to the modified `plotly` object, we can chain calls together with pipes:

```{r}
p <- txhousing %>%
  group_by(city) %>%
  plot_ly(x = ~date, y = ~median) %>%
  add_lines(alpha = 0.2, name = "Texan Cities", hoverinfo = "none") %>%
  filter(city == "Houston") %>%
  add_lines(name = "Houston")
```

## Zooming with context
Want to zoom in without losing context? Try a `rangeslider()`:

```{r}
rangeslider(p)
```

## Any `ggplot` will do...
And just so you don't think we're limited to line charts:

```{r}
p2 <- txhousing %>%
  ggplot(aes(date, median)) + geom_bin2d()
p2 <- ggplotly(p2)
p2
```

Check out [The Plotly Cookbook](https://plotly-book.cpsievert.me/the-plotly-cookbook.html) for more details on specific plotly visualization types ("traces").

# Your Turn!
Find a new data set to practice with and create at least 2 different interactive plots. 