---
title: "uncertainty-vis"
output: html_document
date: "2024-10-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Uncertainty Visualization Demo

In this demo we will review different methods for visualizing uncertainty. We'll use the palmerspenguins data for the demo.

```{r}
library(tidyverse)
library(palmerpenguins)

head(penguins)
```
### Probability Distributions 
First, let's generate a normal probability distribution of mean `bill_length_mm`.

```{r}

bill_mean <- mean(penguins$bill_length_mm, na.rm = TRUE)
bill_sd <- sd(penguins$bill_length_mm, na.rm = TRUE)

# generate normal distribution from mean and sd
bill_norm <- data.frame(rnorm(1000, mean = bill_mean, sd = bill_sd))
names(bill_norm)[1] <- "bill_length_mm"

bill_norm %>% 
  ggplot( aes(x = bill_length_mm)) + 
  geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8)

```

We can also look at the distribution of `bill_length_mm` in our original dataset. 

```{r}

penguins %>% 
  ggplot( aes(x = bill_length_mm)) + 
  geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8)

```


Interesting, maybe our data is bimodal? 

#### Violins

Alternatively, we could use violin plots to show distribution by species.  

```{r}

penguins %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_violin() +
  theme_minimal() 

```


Hmm... maybe only two species are bimodal? 

### Showing Uncertainty for Multiple Points on a Plot 

Now let's look at something a little more interesting. Let's say we want to look at mean bill length by species. We can do this with a bar chart:

```{r}
# data wrangling
df <- penguins %>% 
  group_by(species) %>%
  summarise(mean = mean(bill_length_mm, na.rm = TRUE),
         sd = sd(bill_length_mm, na.rm = TRUE))

# plotting 
df %>%
  ggplot() + 
  geom_bar(aes(x = species, y = mean), stat = "identity", fill="skyblue", alpha=0.5) + 
  theme_minimal()
  
```

#### Error Bars
But a bar chart alone won't show the uncertainty in means. Let's add error bars:

```{r}

df %>%
  ggplot() + 
  geom_bar(aes(x = species, y = mean), stat = "identity", fill = "skyblue", alpha = 0.5) + 
  geom_errorbar(aes(x = species, y = mean, ymin = mean - sd, ymax = mean + sd),
                width = 0.4, color = "orange", alpha = 0.9)

```


Try out `geom_crossbar()`, `geom_linerange()`, and `geom_pointrange()` as well.  

## Challenge 

### HOPs 

Hypothetical Outcome Plots show uncertainty through animation that moves between possible outcomes. For example, a HOP that shows mean `bill_length_mm` by `species`, might show a bar for each `species` that varies in height from `mean - sd` to `mean + sd`, according to probabilities from a normal distribution. 

Your task for the last part of this demo is to **try** to make a HOP version of the bar chart above. You will need to do some data wrangling, and then look into animation. I recommend starting by clearly mapping out your plan for implementing a HOP. Then, generate the data you will need, and finally look at gganimate to help you animate your plot. 

